Dysregulation of Cerebral Hemoglobin Content in Malawian Children with Malaria

### 2021-07-09 figures.Rmd
updated figures with new data

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = "", fig.height = 8, fig.width = 10)

library(tidyverse)
library(janitor)
library(FSA)
library(factoextra)
library(ggrepel)
library(ggpubr)
library(tidymodels)
library(vip)
library(scales)
library(reshape)
library(extrafont)
library(grDevices)

load("glm_model.Rdata")
load("results.Rdata")

@@

### Figure 1: Visual summary of NIRS variables

@@

# hemoglobin concentration
kruskal.test(avg_Hb_conc ~ Status, data = df_hbox)
dunnTest(avg_Hb_conc ~ Status, data = df_hbox, method = "bh")

@@

**Figure 1.** Shown are the variables measured by Oxiplex, plotted by patient status. Cerebral tissue hemoglobin concentration levels were different between cerebral malaria patients and the other two groups, but did not differ between uncomplicated malaria patients and healthy volunteers (Kruskal-Wallis adjusted p-value = 5.00x10^-7^; Dunn Test post-hoc adjusted p-values: CM vs HV = 3.22x10^-6^, CM vs UM = 2.76x10^-4^, HV vs UM = 0.87). Levels of cerebral hemoglobin O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = 0.28).

### Figure 2: Visual summary of α values associated with NIRS variables

@@

# hemoglobin concentration alpha
kruskal.test(Hb_conc_second_alpha ~ Status, data = df_hbox)
dunnTest(Hb_conc_second_alpha ~ Status, data = df_hbox, method = "bh")

# hemoglobin oxygen saturation alpha
kruskal.test(Hb_o2sat_second_alpha ~ Status, data = df_hbox)
dunnTest(Hb_o2sat_second_alpha ~ Status, data = df_hbox, method = "bh")

@@

**Figure 2.** Shown are the alpha values associated with the hemoglobin concentration and oxygen saturation signals measured by Oxiplex. Differences among cerebral tissue hemoglobin concentration alpha values were significant by Kruskal-Wallis (adjusted p-value = 1.57x10^-4^), and differences between all groups were significant by post-hoc Dunn Test (adjusted p-values: CM vs HV = 0.03, CM vs UM = 1.65x10^-4^, HV vs UM = 0.04). Cerebral hemoglobin O~2~ saturation alpha values differed between cerebral malaria patients and the other two patient groups (Kruskal-Wallis adjusted p-value = 2.71x10^-5^; Dunn Test post-hoc adjusted p-values: CM vs HV = 5.18x10^-4^, CM vs UM = 4.44x10^-4^, HV vs UM = 0.23).

@@

#### A. PCA
```{r fig3a}

@@

# select [HbT], [HbT] alpha, and hematocrit
  df_hbox_model <- df_hbox_impute %>% 
    dplyr::select(c(2,4:5,19:22)) %>% 
    mutate(`1/10 * [HbT]α` = `[HbT]α`/10, `[HbT] (nM)` = `[HbT]`/1000) %>% 
    dplyr::select(-c(`[HbT]α`, `[HbT]`)) %>% 
    mutate(Status = as.factor(ifelse(Status == "CM", 1, 0))) %>%

    dplyr::select(-c(`HbO2%`, `HbO2% α`, `Lactate (mmol/L)`))

# resample
set.seed(234)
hbox_boot <- bootstraps(df_hbox_model, times = 1000)

@@

```{r fig3b_plot}

@@

hbox_final %>% 
  tidy(exponentiate = TRUE) %>% 
  mutate(p.value = as.character(p.value)) %>% 
  mutate_if(is.numeric, round, 2) %>% 
  mutate(p.value = as.numeric(p.value)) %>% 
  mutate(p.value = scientific(p.value, digits = 3)) %>% 
  dplyr::rename("Term" = "term", "Odds ratio" = "estimate", "Standard error" = "std.error", 
                "Statistic" = "statistic", "p-value" = "p.value") %>%

@@

### Figure 4: Relationships among hemoglobin concentration, hemoglobin concentration alpha, and hematocrit
```{r fig4}

df_hbox_impute %>% filter(Status != "HV") %>%

  ggplot(aes(x = `[HbT]`, y = `Hematocrit (%)`)) +
  geom_point(aes(color = Status, size = `[HbT]α`), alpha = 0.7) +

  xlab("[HbT] (μM)") +

@@

>>>>

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = "", fig.height = 6, fig.width = 10)

library(tidyverse)
library(DescTools)
library(janitor)
library(FSA)
library(factoextra)
library(ggrepel)
library(ggpubr)
library(tidymodels)
library(vip)
library(scales)
library(reshape)
library(extrafont)
library(grDevices)

load("glm_model.Rdata")
load("results.Rdata")
load("dfa_example.Rdata")
load("dfa_results.Rdata")

@@

### Figure 1: Cerebral tissue hemoglobin concentration and oxygen saturation

@@

# hemoglobin concentration
kruskal.test(avg_Hb_conc ~ Status, data = df_hbox)
DescTools::DunnTest(avg_Hb_conc ~ Status, data = df_hbox, method = "BH")

@@

**Figure 1.** Shown are the variables measured by Oxiplex, plotted by patient status. Cerebral tissue hemoglobin concentration levels were different between cerebral malaria patients and the other two groups, but did not differ between uncomplicated malaria patients and healthy volunteers (Kruskal-Wallis adjusted p-value = 5.0x10^-7^; Dunn Test post-hoc adjusted p-values: CM vs HV = 3.2x10^-6^, CM vs UM = 2.8x10^-4^, HV vs UM = 0.87). Levels of cerebral hemoglobin O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = 0.47).

### Figure 2: Detrended fluctuation analysis results

@@

# hemoglobin concentration alpha
kruskal.test(Hb_conc_second_alpha ~ Status, data = df_hbox)
DescTools::DunnTest(Hb_conc_second_alpha ~ Status, data = df_hbox, method = "BH")

# hemoglobin oxygen saturation alpha
kruskal.test(Hb_o2sat_second_alpha ~ Status, data = df_hbox)
DescTools::DunnTest(Hb_o2sat_second_alpha ~ Status, data = df_hbox, method = "BH")

@@

**Figure 2.** Shown are the alpha values associated with the hemoglobin concentration and oxygen saturation signals measured by Oxiplex. Differences among cerebral tissue hemoglobin concentration alpha values were significant by Kruskal-Wallis (adjusted p-value = 7.16x10^-4^), and differences between uncomplicated malaria patients and the other two patient groups were significant by post-hoc Dunn Test (adjusted p-values: CM vs HV = 0.10, CM vs UM = 4.7x10^-4^, HV vs UM = 0.03). Cerebral hemoglobin O~2~ saturation alpha values differed between cerebral malaria patients and the other two patient groups (Kruskal-Wallis adjusted p-value = 4.2x10^-5^; Dunn Test post-hoc adjusted p-values: CM vs HV = 7.5x10^-4^, CM vs UM = 3.5x10^-4^, HV vs UM = 0.32).

@@

#### A. PCA
```{r fig3a, fig.height = 8}

@@

# select [HbT], [HbT] alpha, and hematocrit
  df_hbox_model <- df_hbox_log_reg %>% 
    dplyr::select(c(Status, Hematocrit, avg_Hb_conc, Hb_conc_second_alpha)) %>% 
    mutate(`1/10_Hb_conc_alpha` = Hb_conc_second_alpha*0.1, avg_Hb_conc_nm = avg_Hb_conc/1000) %>% 
    dplyr::select(-c(Hb_conc_second_alpha, avg_Hb_conc)) %>% 
    mutate(Status = as.factor(ifelse(Status == "CM", 1, 0)))

# resample
set.seed(234)
hbox_boot <- bootstraps(df_hbox_model, strata = Status, times = 1000)

@@

```{r fig3b_plot, fig.height = 8}

@@

hbox_final %>% 
  tidy(exponentiate = TRUE) %>% 
  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2))) %>% 
  mutate(p.value = scientific(p.value, digits = 3),
         std.error = round(std.error, 2),
         statistic = round(statistic, 2)) %>% 
  dplyr::rename("Term" = "term", "Odds ratio" = "estimate", "Standard error" = "std.error", 
                "Statistic" = "statistic", "p-value" = "p.value") %>%

@@

### Figure 4: Relationships among hemoglobin concentration and hematocrit
```{r fig4, fig.height = 8}

df_hbox %>% filter(Status != "HV") %>%

  ggplot(aes(x = avg_Hb_conc, y = Hematocrit)) +
  geom_point(aes(color = Status), alpha = 0.7, size = 3) +

  xlab("Hemoglobin concentration (μM)") +
  ylab("Hematocrit (%)") +

@@

### Supplementary figure 3: Detrended fluctuation analysis

#### A. Pre-processed signal
```{r fig_s3a}

df_thc_cumsum1 %>%

  ggplot(aes(x = Time, y = sg_filt)) +
  geom_line() +

  labs(y = "Hemoglobin concentration (μM)", x = "Time (s)") +
  ggtitle("Pre-processed signal") +
  theme_bw() +
  theme(text = element_text(family = "Arial"),
        axis.text = element_text(size = 20),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 20))

#### B. Cumulative sum
```{r fig_s3b}

df_thc_cumsum1 %>%

  ggplot(aes(x = Time, y = cumsum)) +
  geom_line() +
  geom_vline(xintercept = 1201, lty = 2, color = "black") +
  geom_vline(xintercept = 1210, lty = 2, color = "black") +

  labs(y = "Cumulative sum", x = "Time (s)") +
  ggtitle("Cumulative signal") +
  theme_bw() +
  theme(text = element_text(family = "Arial"),
        axis.text = element_text(size = 20),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 20))

#### C. Window detrend
```{r fig_s3c, fig.show = "hold", out.width = "50%"}

fit <- lm(cumsum ~ Time, data = df_thc_cumsum1 %>% dplyr::filter(Time > 1201 & Time < 1210))

# Window 
df_thc_cumsum1 %>% dplyr::filter(Time > 1201 & Time < 1210) %>%

  ggplot(aes(x = Time, y = cumsum)) +
  geom_line(size = 2) +
  geom_abline(intercept = coef(fit)[1], slope = coef(fit)[2], color = "black", lty = 2, size = 2) +

  labs(y = "Cumulative sum", x = "Time (s)") +
  ggtitle("Cumulative signal") +
  theme_bw() +
  theme(text = element_text(family = "Arial"),
        axis.text = element_text(size = 20),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 20),
        title = element_text(size = 25))

# Detrended
df_thc_cumsum1 %>% dplyr::filter(Time > 1201 & Time < 1210) %>%

  ggplot(aes(x = Time, y = fit$residuals)) +
  geom_line(size = 2) +
  geom_hline(yintercept = 0, color = "black", lty = 2, size = 2) +

  labs(y = "", x = "Time (s)") +
  ggtitle("Detrended cumulative signal") +
  theme_bw() +
  theme(text = element_text(family = "Arial"),
        axis.text = element_text(size = 20),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 20),
        title = element_text(size = 25))

#### D. Results plot
```{r fig_s3d}

# find slope of segmented line
library(segmented)
df1 <- l_dfa_thc[[1]][[1]] %>% mutate(log10_avg_fluct = log10(avg_fluctuation), log10_window_size = log10(window_size))
fit1 <- lm(log10_avg_fluct ~ log10_window_size, data = df1) 
segfit1 <- segmented(fit1)
slopes1 <- coef(segfit1) # the coefficients are the differences in slope in comparison to the previous slope

# first line: 
#y = b0 + b1*x; y = intercept1 + slope1 * x

# second line:
# y = c0 + c1*x; y = intercept2 + slope2 * x

# At the breakpoint (break1), the segments b and c intersect: b0 + b1*x = c0 + c1*x
b0 <- slopes1[[1]]
b1 <- slopes1[[2]]

c1 <- slopes1[[2]] + slopes1[[3]]
break1 <- segfit1$psi[[2]]

# Solve for c0 (intercept of second segment):
c0 <- b0 + b1 * break1 - c1 * break1

df1 %>%

  ggplot(aes(x = log10_window_size, y = log10_avg_fluct)) +
  geom_point(size = 3, shape = 1) +
  geom_abline(intercept = c0, slope = c1, color = "black", lty = 2) +
  annotate("text", x = -0.5, y = 2, label = paste0("alpha = ", round(c1, 2)), size = 8) +

  labs(x = "log10(window length)", y = "log10(average fluctuation)") +
  ggtitle("DFA results") +
